---
title: "careful-whisper_stats"
author: "Tommy Botch"
date: "2024-11-15"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="careful-whisper", plots_name="careful-whisper")
```

# Careful Whisper analyses

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text, or model name) 
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "all-dataset-av-main_careful-whisper_all-results_batch-size-32.csv")

# Load full data (includes both main and shuffled models)
df_all <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('batch_number', 'model_name', 'dataset')
)

# Set factor levels and contrasts on the full dataset before splitting
df_all <- set_factor_levels(
  df=df_all,
  column_name="model_name",
  levels=c("SD", "PA", "AC", "AVC")
)

df_all <- create_column_contrast(
  df=df_all,
  column_name="model_name",
  n_poly=3
)

df_all$is_shuffled <- as.logical(df_all$is_shuffled)

# Main models only (is_shuffled == FALSE)
df_results <- df_all[!df_all$is_shuffled, ]

# Shuffled control frame: SD main rows + PA/AC/AVC shuffled rows
df_shuffled <- rbind(
  df_all[df_all$model_name == "SD" & !df_all$is_shuffled, ],
  df_all[df_all$is_shuffled, ]
)

```

## Analysis 1A - Accuracy of models

### Analysis 1A.1 - Main models

```{r}

accuracy_formula <- 'accuracy ~ model_name + (1|batch_number:dataset)'

model <- lmer(
  accuracy_formula,
  data=df_results,
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

display_model_summary(model)

summary(model)

pairwise_comparisons <- emmeans(model, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model, type=c("pred"))

```

### Analysis 1A.2 - Shuffled (sensory control) models

Tests whether each shuffled model performs at the level of the SD baseline using an LME on a frame containing SD (main) and the shuffled PA/AC/AVC rows. SD serves as the unshuffled reference; pairwise comparisons test each shuffled model against it.

```{r}

model_shuffled <- lmer(
  accuracy_formula,
  data=df_shuffled,
)

tab_model(model_shuffled, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

display_model_summary(model_shuffled)

summary(model_shuffled)

pairwise_comparisons <- emmeans(model_shuffled, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_shuffled, type=c("pred"))

```


## Analysis 1B - Perplexity of models

### Analysis 1B.1 - Main models

```{r}

perplexity_formula <- 'perplexity ~ model_name + (1|batch_number:dataset)'

model <- lmer(
  perplexity_formula,
  data=df_results,
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model)

display_model_summary(model)

pairwise_comparisons <- emmeans(model, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

```

### Analysis 1B.2 - Shuffled (sensory control) models

```{r}

model_shuffled <- lmer(
  perplexity_formula,
  data=df_shuffled,
)

tab_model(model_shuffled, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_shuffled)

display_model_summary(model_shuffled)

pairwise_comparisons <- emmeans(model_shuffled, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_shuffled, type=c("pred"))

```

## Analysis 1C - Information gain of models 

### Analysis 1C.1 - Main models --> greater than SD

```{r}

information_formula <- 'I_total ~ 0 + model_name + (1|batch_number:dataset)'

df_information <- df_results[df_results$model_name != "SD", ]

df_information <- set_factor_levels(
  df_information,
  column_name="model_name",
  levels=c("PA", "AC", "AVC")
)

model_information <- lmer(
  information_formula,
  data=df_information,
)

tab_model(model_information, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_information)

display_model_summary(model_information)

pairwise_comparisons <- emmeans(model_information, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_information, type=c("pred"))

```

### Analysis 1C.2 - Control models (less than SD)

```{r}

information_formula <- 'I_total ~ 0 + model_name + (1|batch_number:dataset)'

df_information_shuffled <- df_shuffled[df_shuffled$model_name != "SD", ]

df_information_shuffled <- set_factor_levels(
  df_information_shuffled,
  column_name="model_name",
  levels=c("PA", "AC", "AVC")
)

model_information <- lmer(
  information_formula,
  data=df_information_shuffled,
)

tab_model(model_information, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_information)

display_model_summary(model_information)

pairwise_comparisons <- emmeans(model_information, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_information, type=c("pred"))

```

### Analysis 1C.3 - Main models trend

```{r}

information_formula <- 'I_total ~ model_name + (1|batch_number:dataset)'

df_information <- df_results[df_results$model_name != "SD", ]

df_information <- set_factor_levels(
  df_information,
  column_name="model_name",
  levels=c("PA", "AC", "AVC")
)

df_information <- create_column_contrast(
  df_information,
  column_name = "model_name",
  n_poly = 2
)

model_information <- lmer(
  information_formula,
  data=df_information,
)

tab_model(model_information, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_information)

display_model_summary(model_information)

pairwise_comparisons <- emmeans(model_information, pairwise ~ model_name)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_information, type=c("pred"))

```

## Analysis 2 - Subset analysis

```{r}

# Load the results file
results_fn <- file.path(results_dir, "all-subsets_careful-whisper_model-comparisons.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('main_model', 'batch_number')
)

# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="main_model", 
  levels=c("SD", "PA", "AC", "AVC")
)

df_results <- create_column_contrast(
  df=df_results, 
  column_name="main_model",
  n_poly=3
)

```

## Analysis 2A - Model accuracy growth curve

```{r}

subsets_formula <- 'main_accuracy ~ main_model * log(true_subset) + (1|batch_number)'

model <- glmer(
  subsets_formula, 
  data=df_results, 
  family=gaussian(link='log'),
  control = glmerControl(optimizer = "Nelder_Mead", optCtrl = list(maxfun = 2e5))
)

summary(model)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ main_model)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model, type='pred', terms=c('true_subset', 'main_model'))

```

### Model perpleixty decay

```{r}

# subject model for accuracy --> test if audio > text
df_cleaned <- remove_outliers(df_results, 'main_perplexity')

df_cleaned <- create_column_contrast(
  df=df_cleaned, 
  column_name="main_model",
  n_poly=3
)

subsets_formula <- 'main_perplexity ~ main_model * log(true_subset) + (1|batch_number)'

model <- glmer(
  subsets_formula, 
  data=df_cleaned, 
  family=gaussian(link='log'),
  control = glmerControl(optimizer ="Nelder_Mead")
)

print (BIC(model))

summary(model)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)
# 
plot_model(model, type='pred', terms=c('true_subset', 'main_model'))

```

### Accuracy subset ratio

```{r}

# set order of variables
df_cleaned <- df_results[df_results$main_model != 'SD',]
df_cleaned <- remove_outliers(df_cleaned, 'subset_accuracy_ratio')

# Set the levels for the modality column
df_cleaned <- set_factor_levels(
  df=df_cleaned,
  column_name="main_model",
  levels=c("PA", "AC", "AVC")
)

df_cleaned <- create_column_contrast(
  df=df_cleaned,
  column_name="main_model",
  n_poly=2
)

subsets_formula <- 'subset_accuracy_ratio ~ main_model * log(true_subset) + (1|batch_number)'

model <- glmer(
  subsets_formula, 
  data=df_cleaned,
  family=gaussian(link='log'),
)
print (AIC(model))

summary(model)
tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)
# 
plot_model(model, type='pred', terms=c('true_subset', 'main_model')) #, show.data=TRUE)

```

### Perplexity subset ratio

```{r}

# set order of variables
df_cleaned <- df_results[df_results$main_model != 'SD',]
df_cleaned <- remove_outliers(df_cleaned, 'subset_accuracy_ratio')

# Set the levels for the modality column
df_cleaned <- set_factor_levels(
  df=df_cleaned,
  column_name="main_model",
  levels=c("PA", "AC", "AVC")
)

df_cleaned <- create_column_contrast(
  df=df_cleaned,
  column_name="main_model",
  n_poly=2
)

subsets_formula <- 'subset_perplexity_ratio ~ main_model * log(true_subset)' # + (1|batch_number)'

model <- glm(
  subsets_formula, 
  data=df_cleaned, 
  family=gaussian(link='log'),
)

print (BIC(model))

summary(model)
tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)
# 
plot_model(model, type='pred', terms=c('true_subset', 'main_model')) #, show.data=TRUE)

```

## Analysis 3 - Information gain

```{r}

# IG values are in the main results CSV (merged per-batch); subset to non-SD main models
df_ig <- df_results[
  df_results$model_name %in% c("PA", "AC", "AVC") &
  !is.na(df_results$delta_I),
  c("batch_number", "dataset", "model_name", "I_total", "I_scr", "delta_I", "delta_I_raw")
]

df_ig$modality <- df_ig$model_name

df_ig <- set_factor_levels(
  df=df_ig,
  column_name="modality",
  levels=c("PA", "AC", "AVC")
)

df_ig <- create_column_contrast(
  df=df_ig,
  column_name="modality",
  n_poly=2
)

```

### Analysis 3A - Structured information gain (delta_I)

```{r}

delta_I_formula <- 'delta_I ~ modality + (1|batch_number:dataset)'

model_ig <- lmer(
  delta_I_formula,
  data=df_ig,
)

tab_model(model_ig, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_ig)

display_model_summary(model_ig)

pairwise_comparisons <- emmeans(model_ig, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_ig, type=c("pred"))

```

### Analysis 3B - Total information gain (I_total)

```{r}

I_total_formula <- 'I_total ~ modality + (1|batch_number:dataset)'

model_itotal <- lmer(
  I_total_formula,
  data=df_ig,
)

tab_model(model_itotal, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

summary(model_itotal)

display_model_summary(model_itotal)

pairwise_comparisons <- emmeans(model_itotal, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

plot_model(model_itotal, type=c("pred"))

```

<!-- ## Load comparisons to humans -->

<!-- ```{r} -->

<!-- normalize <- function(x){(x-min(x))/(max(x)-min(x))} -->

<!-- # Load the results file -->
<!-- results_fn<-glue('{results_dir}/archive-121724/all-task_group-analyzed-behavior_window-size-25_human-prosody-model-distributions-lemmatized.csv') -->
<!-- df_results <- read.csv(results_fn) -->

<!-- df_results <- convert_columns_to_factors(df_results, c('dataset', 'model_name', 'task', 'word_index', 'modality')) -->


<!-- # set order of variables -->
<!-- df_results$model_name <- factor(df_results$model_name, levels = c("ProsodyPrediction", "ProsodyAccess", "ProsodyDeprived")) -->

<!-- # df_results <- remove_outliers(df_results, 'kl_divergence') -->
<!-- ``` -->

<!-- ### Test model KL Divergence -->

<!-- ```{r} -->

<!-- audio_greater_text <- c(1,-1) -->
<!-- contrasts(df_results$modality) <- cbind(audio_greater_text) -->

<!-- # access_greater_deprived <- c(1,-1) -->
<!-- # contrasts(df_results$model_name) <- cbind(access_greater_deprived) -->

<!-- # Test ordered contrast --> audio >  -->
<!-- contrasts(df_results$model_name, how.many=1) <- contr.poly(length(unique(df_results$model_name))) -->

<!-- kl_formula = 'kl_divergence ~ modality * model_name * prominence_mean + (1|word_index:task) + (1|dataset)' -->

<!-- model <- lmer(kl_formula, data=df_results, REML=TRUE) -->

<!-- tab_model(model) -->

<!-- summary(model) -->
<!-- plot_model(model, type='pred', terms=c('modality', 'model_name')) -->

<!-- # ggplot(df_results, aes(x=sqrt(prominence_mean))) + geom_density() -->
<!-- ``` -->

<!-- ```{r} -->

<!-- check_predictions(model) -->

<!-- testDispersion(model) -->
<!-- simulationOutput <- simulateResiduals(model, plot = F) -->
<!-- plot(simulationOutput) -->
<!-- ``` -->