---
title: "results04_human-llm_accuracy-stats"
author: "Tommy Botch"
date: "2025-10-02"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral", plots_name="human-llm-comparison")
```

# Human-LLM accuracy analyses

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text, or model name) 
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "all-task_group-analyzed-behavior_window-size-00025_human-model-lemmatized.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('modality', 'word_index', 'task', "Clause_Pos", "Clause_ID", "POS_spaCy", "ground_truth")
)

# ADD A COLUMN THAT GROUPS HUMANS/MODELS
df_results <- df_results %>%
  mutate(model_type = case_when(
    modality %in% c("video") ~ "video",
    modality %in% c("audio") ~ "audio",
    modality %in% c("text") ~ "text",
    modality %in% c('roberta', 'electra', 'xlm-prophetnet') ~ "mlm",
    TRUE ~ "clm"  # Default case
  ))

# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="model_type",
  levels=c("video", "audio", "text", "clm", "mlm")
)

df_results <- create_column_contrast(
  df=df_results, 
  column_name="model_type",
  n_poly=2
)

```

## Analysis 1A - Binary accuracy

We first investigate whether humans exhibit an advantage of binary accuracy (0/1) over LLMs based on the modality of the stimulus. We use a logistic mixed-effects model to classify accuracy based on modality.

```{r}

binary_accuracy_formula <- 'accuracy ~ model_type + (1|modality) + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- glmer(
  binary_accuracy_formula, 
  data=df_results, 
  family=binomial(link="logit")
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)

```
## Analysis 1B - Task-level comparisons

```{r}

binary_accuracy_formula <- 'accuracy ~ model_type + (1|modality) + (1|word_index)'

```

### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_results, 
  formula=binary_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_results, 
  formula=binary_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_results, 
  formula=binary_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

## Analysis 2A - Continuous accuracy

```{r}

continuous_accuracy_formula <- 'fasttext_top_word_accuracy ~ model_type + (1|modality) + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- lmer(
  continuous_accuracy_formula, 
  data=df_results, 
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)

```

## Analysis 2B - Task-level comparisons

```{r}

continuous_accuracy_formula <- 'fasttext_top_word_accuracy ~ model_type + (1|modality) + (1|word_index)'

```

### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_results, 
  formula=continuous_accuracy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_results, 
  formula=continuous_accuracy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_results, 
  formula=continuous_accuracy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ model_type)
display_pairwise_comparisons(pairwise_comparisons)
```

## Analysis 2C - Continuous accuracy by quadrant

```{r}

# Create accuracy-entropy column removing any missing
n_before <- nrow(df_results)

df_accuracy_entropy <- df_results %>%
  mutate(
    accuracy_group = na_if(accuracy_group, ""),
    entropy_group  = na_if(entropy_group, "")
  ) %>%
  drop_na(accuracy_group, entropy_group) %>%
  mutate(accuracy_entropy_group = paste(accuracy_group, entropy_group, sep = "-"),
         accuracy_entropy_group = factor(accuracy_entropy_group))

n_after <- nrow(df_accuracy_entropy)
cat("Rows dropped:", n_before - n_after, "\n")

# Make factor
df_accuracy_entropy <- convert_columns_to_factors(
  df_accuracy_entropy,
  c("accuracy_entropy_group")
)

# Set the levels for the modality column
df_accuracy_entropy <- set_factor_levels(
  df=df_accuracy_entropy,
  column_name="accuracy_entropy_group",
  levels=c("high-low", "high-high", "low-low", "low-high")
)

unique(df_accuracy_entropy$accuracy_entropy_group)

df_accuracy_entropy <- create_column_contrast(
  df=df_accuracy_entropy, 
  column_name="accuracy_entropy_group",
  n_poly=1
)
```

Now run the dataframe through

```{r}

formula <- "fasttext_top_word_accuracy ~ accuracy_entropy_group * model_type + (1|modality) + (1|word_index)"

# Fit the logistic mixed-effects model
model <- lmer(
  formula, 
  data=df_accuracy_entropy, 
)


tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Estimated marginal means for model_type, conditional on accuracy_entropy_group = "high-low"
emm_highlow <- emmeans(model, pairwise ~ model_type | accuracy_entropy_group)


emm_highlow

# Plot predicted values for interaction
plot_model(model, type = "pred", terms = c("accuracy_entropy_group", "model_type"))
```

```{r}

# Get estimated marginal means for the interaction
emm <- emmeans(model, ~ accuracy_entropy_group * model_type)

# Convert to data frame for plotting
emm_df <- as.data.frame(emm)

# Line plot
ggplot(emm_df, aes(
  x = accuracy_entropy_group, 
  y = emmean, 
  color = model_type, 
  group = model_type
)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  # geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
  #               width = 0.1) +
  theme_minimal() +
  labs(
    x = "Accuracy × Entropy group",
    y = "Estimated probability of top word accuracy",
    color = "Model type"
  )

```

## Analysis 3A - Context window continuous accuracy

Load results again

```{r}
results_fn <- file.path(results_dir, "all-task_group-analyzed-behavior_all-window-sizes_human-model-lemmatized.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('modality', 'word_index', 'task')
)

# ADD A COLUMN THAT GROUPS HUMANS/MODELS
df_results <- df_results %>%
  mutate(model_type = case_when(
    modality %in% c("video") ~ "video",
    modality %in% c("audio") ~ "audio",
    modality %in% c("text") ~ "text",
    modality %in% c('roberta', 'electra', 'xlm-prophetnet') ~ "mlm",
    TRUE ~ "clm"  # Default case
  ))

# Drop MLM rows
df_results <- df_results %>%
  filter(model_type != "mlm")


# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="model_type",
  levels=c("video", "audio", "text", "clm") #, "mlm")
)

df_results <- create_column_contrast(
  df=df_results,
  column_name="model_type",
  n_poly=3
)
# 
# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="modality",
  levels=c("video", "audio", "text", "llama3.1-70B", "qwen3-32B", "mistral", "gpt2-xl", "gpt-neo-x", "gemma3-1b-pt", "llama2", "bloom", "gpt2", "roberta", "electra", "xlm-prophetnet")
)


# df_results <- create_column_contrast(
#   df=df_results,
#   column_name="modality",
#   n_poly=4
# )

unique(df_results$modality)

```

Run analysis

```{r}

df_results$log_window <- log(df_results$window_size)

continuous_accuracy_formula <- 'fasttext_top_word_accuracy ~ model_type * log_window + (1|modality) + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- lmer(
  continuous_accuracy_formula, 
  data=df_results, 
  # family=gaussian(link="log")
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Step 1: Define a grid of window sizes
window_vals <- sort(unique(df_results$log_window))

# Step 2: Compute predicted marginal means for each model_type at each window size
emm <- emmeans(model, pairwise ~ model_type | log_window, at = list(log_window = window_vals))

emm
# Step 3: Convert emmeans output to a data frame
pred_df <- as.data.frame(emm)
AIC(model)

# Line plot
ggplot(pred_df, aes(
  x = log_window, 
  y = emmean, 
  color = model_type, 
  group = model_type
)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  # geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
  #               width = 0.1) +
  theme_minimal() +
  labs(
    x = "Accuracy × Entropy group",
    y = "Estimated probability of top word accuracy",
    color = "Model type"
  )


```