---
title: "supp02_exploratory-features"
author: "Tommy Botch"
date: "2024-11-15"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral", plots_name="behavioral")
```

# Population-level analysis 

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text) 
- **ground_truth:** ground-truth word to be predicted
- **top_pred:** most predicted word across the population
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "exploratory-features-analysis.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('task', 'word_index', 'ground_truth', "POS_spaCy", "Clause_Pos", "Clause_ID")
)

# Drop multiple columns
colnames(df_results)
```

### Run LME

```{r}

formula <- "improvement_slope ~ Frequency + Valence + Arousal + Concreteness + Length + sentiment_score + (1|task) + (1|POS_spaCy) + (1|Clause_Pos:Clause_ID)"

model <- lmer(
  formula=formula,
  data=df_results,
)

tab_model(model)
```
