---
title: "supp01_prosody-syntax"
author: "Tommy Botch"
date: "2024-11-15"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral", plots_name="behavioral")
```

# Population-level analysis 

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text) 
- **ground_truth:** ground-truth word to be predicted
- **top_pred:** most predicted word across the population
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

tasks <- c("black", "wheretheressmoke", "howtodraw")

df_all <- map_dfr(tasks, function(task) {
  prosody_syntax_fn <- file.path(
    stim_dir,
    task,
    paste0(task, "_transcript-selected_prosody-syntax.csv")
  )
  
  df <- load_and_prepare_data(
    file_path = prosody_syntax_fn,
    factor_columns = c("POS", "POS_spaCy", "Dep", "Head", "Clause_ID", "Clause_Pos", "word")
  )
  
  df$task <- task
  return(df)
})

colnames(df_all)

# Turn task into a factor
df_all$task <- as.factor(df_all$task)

```
Collapse categories a bit

```{r}

df_all <- df_all %>%
  mutate(POS_group = case_when(
    POS_spaCy %in% c("NOUN","PROPN","PRON","NUM") ~ "Nominal",
    POS_spaCy %in% c("VERB","AUX") ~ "Verb",
    POS_spaCy %in% c("ADJ","ADV") ~ "Modifier",
    # POS_spaCy %in% c("CCONJ","SCONJ") ~ "Function"
  )) %>%
  mutate(POS_group = factor(POS_group))

# Set factor levels and reference
df_all$POS_group <- factor(
  df_all$POS_group,
  levels = c("Nominal", "Verb", "Modifier")  # ordered as desired
)

df_all <- na.omit(df_all)

# Remove non-NWP candidates
df_sub <- subset(df_all, df_all$NWP_Candidate == "True")

# Make sure POS is a factor with all levels kept
df_sub$POS_spaCy <- droplevels(df_sub$POS_spaCy)

```


## Analysis 1 -- POS Correlation w/ Prosody

```{r}

levels(df_all$POS_spaCy)

formula <- "prominence_1 ~ -1 + POS_group + (1|Clause_Pos) + (1|word) + (1|task)"

model <- lmer(formula, data=df_sub)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

plot_model(model, type="pred", terms=c("POS_group"))

preds <- data.frame(get_model_data(model, type='pred', terms = c("POS_group")))

preds

ggplot(preds, aes(x=x, y=predicted)) +
  geom_point(
    data = df_sub,
    aes(x=POS_group, y=prominence),
    stroke=1,
    alpha=0.035
  ) +
  # Add predicted mean points
  geom_point(
    data = preds,
    aes(x=x, y=predicted),
    color="red",
    size=3
  ) +
  # Add error bars for predictions
  geom_errorbar(
    data = preds,
    aes(x=x, ymin=conf.low, ymax=conf.high),
    width=0.2,
    color="red"
  ) +
  theme(
    aspect.ratio = 3/4,
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  scale_color_manual(values=colors) +
  scale_fill_manual(values=colors) +
  labs(
    x = "Part of Speech",
    y = "Prosodic Prominence",
    title = "Relationship of POS and Prosody"
  )

```
## Analysis 2 - Lagged prosody

Examine whether prosody at the word before a given word predicts the upcoming part of speech

## Binary models

```{r}

library(glue)

# ===================================================================
# APPROACH 1: SEPARATE BINARY MODELS (Most interpretable)
# ===================================================================

# Test each category vs. all others
binary_models <- list()
categories <- levels(df_sub$POS_group)

prominence_entry <- "prominence_9"

for(cat in categories) {
  # Create binary outcome
  df_sub[[paste0("is_", cat)]] <- ifelse(df_sub$POS_group == cat, 1, 0)
  
  # Fit binary model with tryCatch
  model <- tryCatch(
    {
      glmer(
        as.formula(glue("is_{cat} ~ {prominence_entry} + (1|Clause_Pos:Clause_ID) + (1|task)")),
        data = df_sub,
        family = binomial()
      )
    },
    error = function(e) {
      message("Model failed for category: ", cat, " -- skipping")
      return(NULL)  # return NULL if the model fails
    }
  )
  
  # Only save the model if it succeeded
  if (!is.null(model)) {
    binary_models[[cat]] <- model
  }
}

# Extract and compare effects including p-values
binary_effects <- map_dfr(names(binary_models), ~ {
  model <- binary_models[[.x]]
  
  # Coefficient estimate for prominence_lag1
  est <- fixef(model)[prominence_entry]
  
  # Standard error
  se <- sqrt(diag(vcov(model)))[prominence_entry]
  
  # 95% CI using Wald approximation
  ci_low <- est - 1.96 * se
  ci_high <- est + 1.96 * se
  
  # Wald z-test for significance
  z_value <- est / se
  p_value <- 2 * (1 - pnorm(abs(z_value)))
  
  data.frame(
    category = .x,
    estimate = est,
    conf.low = ci_low,
    conf.high = ci_high,
    z = z_value,
    p.value = p_value
  )
})

print("Binary model effects with significance:")
print(binary_effects)


# Plot binary effects
ggplot(binary_effects, aes(x = category, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Binary Model Effects: Each Category vs. Others",
       x = "POS Category", y = "Log-odds of being in category") +
  theme_minimal()

```

## Multinomial Classificiation

```{r}
library(mclogit)

# -------------------------------
# 1. Set predictor variable here
# -------------------------------
predictor_var <- "prominence_mean_words5"  # <-- change this to any numeric predictor

# -------------------------------
# 2. Fit the multinomial model
# -------------------------------
modele <- mblogit(
  formula = as.formula(paste("POS_group ~ ", predictor_var)),
  random = list(~1|task, ~1|Clause_Pos),
  data = df_sub,
  method = "PQL",
  estimator = "ML"
)

tab_model(modele)
# -------------------------------
# 3. Create prediction data frame
# -------------------------------
preds <- predict(modele, type = "response")
pred_df <- as.data.frame(preds)

# Attach the predictor column dynamically
pred_df[[predictor_var]] <- df_sub[[predictor_var]]

# Pivot to long format
pred_long <- pred_df %>%
  pivot_longer(
    cols = -all_of(predictor_var),
    names_to = "POS_group",
    values_to = "probs"
  )

# -------------------------------
# 4. Aggregate predictions for smooth plotting
# -------------------------------
# Bin the predictor to reduce noise
pred_summary <- pred_long %>%
  mutate(bin = cut(.data[[predictor_var]], breaks = 10)) %>%
  group_by(bin, POS_group) %>%
  summarize(
    mean_prob = mean(probs),
    se_prob = sd(probs)/sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    conf.low = mean_prob - 1.96*se_prob,
    conf.high = mean_prob + 1.96*se_prob
  )

# Baseline (chance) probability per category
baseline <- pred_long %>%
  group_by(POS_group) %>%
  summarize(baseline_prob = mean(probs), .groups = "drop")


```

### Plot classification probability based on prominence

```{r}

# -------------------------------
# 5. Plot
# -------------------------------
ggplot(pred_summary, aes(x = as.numeric(bin), y = mean_prob, color = POS_group, group = POS_group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = POS_group), alpha = 0.2, color = NA) +
  geom_hline(
    data = baseline,
    aes(yintercept = baseline_prob, color = POS_group),
    linetype = "dashed", size = 0.8
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = 1:10,
    labels = round(seq(min(df_sub[[predictor_var]]), max(df_sub[[predictor_var]]), length.out = 10), 2)
  ) +
  labs(
    x = predictor_var,
    y = "Predicted Probability",
    title = paste("Predicted Probabilities vs", predictor_var)
  ) +
  theme(legend.title = element_blank())

```
### Testing different appraoch

```{r}
pos_levels <- levels(df_sub$POS_group)
n <- nrow(df_sub)

# -----------------------------
# 1. Marginal frequency baseline (no sampling needed)
# -----------------------------
marginal_probs <- prop.table(table(df_sub$POS_group))
# Predict most frequent class for everyone
most_frequent_class <- names(which.max(marginal_probs))
pred_marginal <- factor(rep(most_frequent_class, n), levels = pos_levels)

# pred_marginal <- sample(names(marginal_probs), 
#                                size = n, 
#                                replace = TRUE, 
#                                prob = marginal_probs)

# Create confusion matrix
conf_marginal <- table(Predicted = pred_marginal, Actual = df_sub$POS_group)
conf_marginal_df <- as.data.frame(conf_marginal) %>%
  mutate(Freq_norm = Freq / n, Type = "Marginal Baseline")

# -----------------------------
# 2. Intercept-only model baseline
# -----------------------------

# Fit intercept-only model
intercept_model <- mblogit(
  POS_group ~ 1,
  random = list(~1|task, ~1|Clause_Pos),
  data = df_sub,
  method = "PQL",
  estimator = "ML"
)

# Get predictions from intercept-only model
pred_intercept_probs <- predict(intercept_model, type = "response")
pred_intercept_class <- factor(colnames(pred_intercept_probs)[max.col(pred_intercept_probs)], 
                              levels = pos_levels)

conf_intercept <- table(Predicted = pred_intercept_class, Actual = df_sub$POS_group)
conf_intercept_df <- as.data.frame(conf_intercept) %>%
  mutate(Freq_norm = Freq / n, Type = "Intercept-only")

# -----------------------------
# 3. Full model predictions (recalculate the original model)
# -----------------------------
# This should match your original model exactly
full_model <- mblogit(
  formula = as.formula(paste("POS_group ~ ", predictor_var)),
  random = list(~1|task, ~1|Clause_Pos),
  data = df_sub,
  method = "PQL",
  estimator = "ML"
)

# Get predictions from the full model
pred_full_probs <- predict(full_model, type = "response")
pred_class <- factor(colnames(pred_full_probs)[max.col(pred_full_probs)], levels = pos_levels)

conf_full <- table(Predicted = pred_class, Actual = df_sub$POS_group)
conf_full_df <- as.data.frame(conf_full) %>%
  mutate(Freq_norm = Freq / n, Type = "Full Model")

# -----------------------------
# 4. Calculate performance metrics
# -----------------------------
calculate_metrics <- function(conf_matrix, model_name) {
  # Overall accuracy
  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
  
  # Per-class metrics
  precision <- diag(conf_matrix) / rowSums(conf_matrix)
  recall <- diag(conf_matrix) / colSums(conf_matrix)
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  # Handle NaN values
  precision[is.nan(precision)] <- 0
  recall[is.nan(recall)] <- 0
  f1[is.nan(f1)] <- 0
  
  data.frame(
    Model = model_name,
    Accuracy = accuracy,
    Mean_Precision = mean(precision, na.rm = TRUE),
    Mean_Recall = mean(recall, na.rm = TRUE),
    Mean_F1 = mean(f1, na.rm = TRUE)
  )
}

# Calculate metrics for all models
metrics_marginal <- calculate_metrics(conf_marginal, "Marginal Baseline")
metrics_intercept <- calculate_metrics(conf_intercept, "Intercept-only")
metrics_full <- calculate_metrics(conf_full, "Full Model")

performance_summary <- bind_rows(metrics_marginal, metrics_intercept, metrics_full)
print("Model Performance Comparison:")
print(performance_summary)

# -----------------------------
# 5. Combine confusion matrices for visualization
# -----------------------------
conf_combined <- bind_rows(conf_marginal_df, conf_intercept_df, conf_full_df) %>%
  mutate(Type = factor(Type, levels = c("Marginal Baseline", "Intercept-only", "Full Model")))

# -----------------------------
# 6. Enhanced visualization
# -----------------------------
p1 <- ggplot(conf_combined, aes(x = Actual, y = Predicted, fill = Freq_norm)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = paste0(Freq, "\n(", round(Freq_norm, 3), ")")), 
            color = "white", size = 3) +
  facet_wrap(~Type) +
  scale_fill_gradient(low = "steelblue", high = "darkred", name = "Proportion") +
  scale_y_discrete(limits = rev(pos_levels)) +
  coord_fixed() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Model Performance Comparison w/ Random Effects: Confusion Matrices",
       subtitle = "Numbers show count (proportion)",
       x = "Actual POS",
       y = "Predicted POS")

# Performance metrics visualization
p2 <- performance_summary %>%
  select(-Accuracy) %>%
  tidyr::pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Model Performance Metrics",
       y = "Score",
       x = "Model Type")

# Print plots
print(p1)
print(p2)

# -----------------------------
# 7. Statistical significance test (McNemar's test for paired comparisons)
# -----------------------------
# Compare full model vs intercept-only
mcnemar_test <- function(conf1, conf2) {
  # Create contingency table for McNemar's test
  correct1 <- diag(conf1)
  correct2 <- diag(conf2)
  
  # This is a simplified approach - for proper McNemar's test with multinomial,
  # you might want to use more sophisticated methods
  overall_acc1 <- sum(correct1) / sum(conf1)
  overall_acc2 <- sum(correct2) / sum(conf2)
  
  cat("Accuracy comparison:\n")
  cat("Full Model:", round(overall_acc1, 4), "\n")
  cat("Intercept-only:", round(overall_acc2, 4), "\n")
  cat("Improvement:", round(overall_acc1 - overall_acc2, 4), "\n")
}

mcnemar_test(conf_full, conf_intercept)

# -----------------------------
# 8. Class-wise improvement analysis
# -----------------------------
class_improvement <- data.frame(
  POS_group = pos_levels,
  Marginal_Accuracy = diag(conf_marginal) / colSums(conf_marginal),
  Intercept_Accuracy = diag(conf_intercept) / colSums(conf_intercept),
  Full_Accuracy = diag(conf_full) / colSums(conf_full)
) %>%
  mutate(
    Improvement_over_Marginal = Full_Accuracy - Marginal_Accuracy,
    Improvement_over_Intercept = Full_Accuracy - Intercept_Accuracy
  )

print("Class-wise Performance:")
print(class_improvement)

```


### Plot confusion matrices

```{r}


pos_levels <- levels(df_sub$POS_group)
n <- nrow(df_sub)

# -----------------------------
# 1. Native baseline: sample according to marginal frequencies
# -----------------------------
native_probs <- prop.table(table(df_sub$POS_group))

set.seed(123)
native_class_sample <- sample(names(native_probs),
                              size = n,
                              replace = TRUE,
                              prob = native_probs)

conf_native <- table(Predicted = native_class_sample,
                     Actual = df_sub$POS_group) %>%
  as.data.frame() %>%
  rename(Count = Freq) %>%
  mutate(Freq = Count / n,
         Type = "Native") %>%
  select(Predicted, Actual, Freq, Type)

# -----------------------------
# 2. Intercept-only baseline
# -----------------------------
coefs <- coef(modele)
intercepts_vec <- coefs[grep("\\(Intercept\\)", names(coefs))]
names(intercepts_vec) <- gsub("~\\(Intercept\\)", "", names(intercepts_vec))

exp_int <- exp(intercepts_vec)
p_nominal <- 1 / (1 + sum(exp_int))
p_other <- exp_int / (1 + sum(exp_int))
baseline_probs_vec <- c(Nominal = p_nominal, p_other)

set.seed(123)
baseline_class_prob <- sample(names(baseline_probs_vec),
                              size = n,
                              replace = TRUE,
                              prob = baseline_probs_vec)

conf_baseline <- table(Predicted = baseline_class_prob,
                       Actual = df_sub$POS_group) %>%
  as.data.frame() %>%
  rename(Count = Freq) %>%
  mutate(Freq = Count / n,
         Type = "Intercept-only") %>%
  select(Predicted, Actual, Freq, Type)

# -----------------------------
# 3. Observed predictions
# -----------------------------

pred_class <- colnames(preds)[max.col(preds)]
pred_class <- factor(pred_class, levels = pos_levels)

conf_observed <- table(Predicted = pred_class,
                       Actual = df_sub$POS_group) %>%
  as.data.frame() %>%
  rename(Count = Freq) %>%
  mutate(Freq = Count / n,
         Type = "Observed") %>%
  select(Predicted, Actual, Freq, Type)

# -----------------------------
# 4. Combine all three and set order
# -----------------------------
conf_combined <- bind_rows(conf_native, conf_baseline, conf_observed) %>%
  mutate(Type = factor(Type, levels = c("Native", "Intercept-only", "Observed")))

# -----------------------------
# 5. Plot all as frequency confusion matrices
# -----------------------------
ggplot(conf_combined, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), color = "white") +
  facet_wrap(~Type) +
  scale_fill_gradient(low = "steelblue", high = "darkred") +
  scale_y_discrete(limits = rev(pos_levels)) +
  coord_fixed() +  # ensures squares
  theme_minimal() +
  labs(title = "POS Distribution Comparison (Sampled Frequencies)",
       x = "Actual POS",
       y = "Predicted POS",
       fill = "Frequency")
```
