---
title: "results03_population-probability-stats"
author: "Tommy Botch"
date: "2024-11-15"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral", plots_name="behavioral")
```

# Population-level analysis 

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text) 
- **ground_truth:** ground-truth word to be predicted
- **top_pred:** most predicted word across the population
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "all-task_group-analyzed-behavior_human-lemmatized.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('modality', 'ground_truth', 'top_pred', 'word_index','task',"POS_spaCy", "Clause_Pos", "Clause_ID")
)

# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="modality",
  levels=c("video", "audio", "text")
)

df_results <- create_column_contrast(
  df=df_results, 
  column_name="modality",
)
```

## Analysis 1A - Predictability

We first investigate whether predictability (percentage of subjects predicting the correct word) varies based on the modality of the stimulus. We use a linear mixed-effects model with a log-link function, given that predictability is log distributed (e.g., floor is most common).

```{r}

predictability_formula <- 'predictability ~ modality + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- glmer(predictability_formula, data=df_results, family=gaussian(link='log'))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

## Analysis 1B - Predictability interaction with prosody

```{r}

predictability_formula <- 'predictability ~ modality * prominence_mean_words5 + (1|word_index:task)'

# Fit the mixed-effects model
model <- glmer(predictability_formula, data=df_results, family=gaussian(link='log'))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```
### Plot the interaction between predictability and prosody

```{r}

plot_model(model, type='pred', terms=c('prominence_mean_words5', 'modality'))

```

## Analysis 1C - Task-level comparisons

```{r}

predictability_formula <- 'predictability ~ modality + (1|word_index)'

```

Now conduct the comparisons for each task.

### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_results, 
  formula=predictability_formula,
  family=gaussian(link="log")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_results, 
  formula=predictability_formula,
  family=gaussian(link="log")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_results, 
  formula=predictability_formula,
  family=gaussian(link="log")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

## Analysis 2A - Entropy

We next look at whether entropy of human prediction distributions varies based on condition.  

```{r}

entropy_formula <- 'entropy ~ modality + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- lmer(entropy_formula, data = df_results)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)


# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```


## Analysis 2B - Entropy interaction with prosody

Include prosody (prominence_mean) within the model as an interaction effect.

```{r}

entropy_formula <- 'entropy ~ modality * prominence_mean_words5 + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- lmer(entropy_formula, data = df_results)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Plot the interaction between prominence and entropy

```{r}

preds <- data.frame(get_model_data(model, type='pred', terms = c("prominence_mean_words5", "modality")))

colors <- c('#F068A1', '#82C564', '#F7CD84')

ggplot(preds, aes(x=x, y=predicted, color=group)) +
  geom_point(
    data = df_results,
    aes(x=prominence_mean_words5, y=entropy, color=modality),
    stroke=1,
    alpha=0.1) +
  geom_line(size=1) +
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill=group), alpha=0.3, color=NA) +
  theme(
    aspect.ratio = 5/4,
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  scale_color_manual(values=colors) +
  scale_fill_manual(values=colors) +
  labs(
    x = "Prosodic Prominence",
    y = "Entropy",
    title = "Prosody reduces entropy of prediction distributions")

ggsave(file.path(plots_dir, 'human_entropy-prosody-interaction_new.pdf'), device = "pdf")
```





## Analysis 2C - Task-level comparisons

```{r}

entropy_formula <- 'entropy ~ modality + (1|word_index)'

```

Now conduct the comparisons for each task.

### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_results, 
  formula=entropy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_results, 
  formula=entropy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_results, 
  formula=entropy_formula,
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

## Supplementary Analysis - Leakage filtering 

We now want to ensure that observed differences were not driven by confounding factors such as "leakage" of auditory input. We therefore filter for any instance where 

```{r}

df_results_filtered <- subset(df_results, !df_results$filter_for_leakage)

```

### Supplementary Analysis 1A - Predictability

```{r}

predictability_formula <- 'predictability ~ modality + (1|word_index:task)'

# Fit the mixed-effects model
model <- glmer(predictability_formula, data=df_results_filtered, family=gaussian(link='log'))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```



### Supplementary Analysis 2A - Entropy


```{r}

entropy_formula <- 'entropy ~ modality + (1|word_index:task)'

# Fit the logistic mixed-effects model
model <- lmer(entropy_formula, data = df_results_filtered)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

