---
title: "results01_subject-accuracy-stats"
author: "Tommy Botch"
date: "2025-10-02"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral", plots_name="behavioral")
```

# Participant-level analysis 

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text) 
- **prolific_id:** the unique identifier of each prolific participant
- **subject:** the subject number yoked to presentation order (repeated across modalities)
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "all-task_subject-behavior_lemmatized.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c("modality", "prolific_id", "subject", "word_index", "task", "POS_spaCy", "Clause_ID", "Clause_Pos")
)

# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="modality",
  levels=c("video", "audio", "text")
)

df_results <- create_column_contrast(
  df=df_results, 
  column_name="modality"
)

colnames(df_results)
```

## Analysis 1A - Average accuracy of individual participants

We first test whether there is an advantage of modality at the level of individual participants. To test this hypothesis, we first calculate the average accuracy within a given participant across trials.

```{r}

# Calculate the average accuracy per subject within modality and task
df_avg <- df_results %>%
  group_by(subject, modality, task) %>%
  summarise(avg_accuracy = mean(accuracy), .groups = 'drop')

```

We then predict the average accuracy of participants based on their modality, while controlling for random effect of presentation order (e.g., the identifier used for a subject within a task).

```{r}

average_accuracy_formula <- 'avg_accuracy ~ modality + (1|subject:task)'

# Fit the logistic mixed-effects model
model <- lmer(average_accuracy_formula, data=df_avg)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Analysis 1B - Task-level comparisons

We are also interested in understanding how these effects exist at the level of individual tasks. Since task is a random effect in previous models, we therefore run separate models for each tasks.

We first setup our formula to evaluate our model -- we discard task are random-effect as we are inherently testing at the task-level. 

```{r}

task_avg_accuracy_formula <- 'avg_accuracy ~ modality + (1|subject)'

```

We now run this model separately for each task, enabling us to conduct pairwise comparisons of conditions within task.

#### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_avg, 
  formula=task_avg_accuracy_formula
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

#### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_avg, 
  formula=task_avg_accuracy_formula
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

#### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_avg, 
  formula=task_avg_accuracy_formula
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```


## Analysis 2A - Accuracy of individual participants

We can also investigate whether there are differences in the raw accuracy of participants within a condition (modality) while not collapsing across trials (e.g., avoiding averaging across participants).

To do this, we construct a logistic mixed-effects model that predicts binary accuracy (0/1) based on modality. We control for the random effect of trial (word_index) within task, as well as the random effect of presentation order (subject) within modality and task.

```{r}

formula <- 'accuracy ~ modality + (1|word_index:task) + (1|subject:modality:task)'

# Fit the logistic mixed-effects model
model <- glmer(formula, 
               data = df_results, 
               family = binomial(link = "logit"))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

summary(model)
```

## Analysis 2B - Interaction of accuracy with prosody (individual participants)

We also see if prosody (prominence) over the words leading up to the to-be-predicted word relate to the differences in condition (modality). This analysis can only be conducted when we consider individual trials (word_index), as we have a prominence value associated with each word.

```{r}

subject_accuracy_formula <- 'accuracy ~ modality * prominence_mean_words5 + (1|word_index:task) + (1|subject:modality:task)'

# Fit the logistic mixed-effects model
model <- glmer(subject_accuracy_formula, 
               data = df_results, 
               family = binomial(link = "logit"))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE)
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

## Analysis 2C - Task-level comparisons

Similar to before, we investigate the effects within-task to conduct pairwise comparisons between conditions. We first set up the formula for the comparison. 

```{r}

task_accuracy_formula <- 'accuracy ~ modality + (1|word_index) + (1|subject:modality)'

```

Now conduct the comparisons for each task.

#### Story = Black

```{r}

model <- lmer_task(
  task_name='black', 
  df=df_results, 
  formula=task_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)
```

#### Story = Smoke

```{r}

model <- lmer_task(
  task_name='wheretheressmoke', 
  df=df_results, 
  formula=task_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

#### Story = Draw

```{r}

model <- lmer_task(
  task_name='howtodraw', 
  df=df_results, 
  formula=task_accuracy_formula,
  family=binomial(link="logit")
)

# Output the results
tab_model(model, show.se = TRUE, show.stat = TRUE, show.r2 = TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```


## Supplementary Analysis - Leakage filtering 

We now want to ensure that observed differences were not driven by confounding factors such as "leakage" of auditory input. We therefore filter for any instance where 

```{r}

df_results_filtered <- subset(df_results, !df_results$filter_for_leakage)

```

### Supplementary Analysis 1 - Average accuracy of individual participants

We rerun our analyses of average accuracy after filtering for leakage. We first average trials within subject to compute the average accuracy. 

```{r}

# Calculate the average accuracy per subject within modality and task
df_avg <- df_results_filtered %>%
  group_by(subject, modality, task) %>%
  summarise(avg_accuracy = mean(accuracy), .groups = 'drop')

```

We then predict the average accuracy of participants based on their modality, while controlling for random effect of presentation order (e.g., the identifier used for a subject within a task).

```{r}

average_accuracy_formula <- 'avg_accuracy ~ modality + (1|subject:task)'

# Fit the logistic mixed-effects model
model <- lmer(average_accuracy_formula, data=df_avg)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Supplementary Analysis 2 - Accuracy across individuals

We repeat our analysis of raw accuracy across subjects within modality, after filtering for leakage.

```{r}

subject_accuracy_formula <- 'accuracy ~ modality + (1|word_index:task) + (1|subject:modality:task)'

# Fit the logistic mixed-effects model
model <- glmer(subject_accuracy_formula, 
               data = df_results_filtered, 
               family = binomial(link = "logit"))

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```