---
title: "Shortened Removed: results05_human-llm_accuracy-difference-stats"
author: "Tommy Botch"
date: "2025-10-02"
output: html_document
---

# Setup environment

## Load packages and setup paths for analysis

Set up directories - since we're doing this locally (and mounting the server), 
we're going to use a different base_dir

```{r setup, include=FALSE}

# Source the utils.R file to load the functions
source("../utils/utils.R")

# Load required and additional packages
install_and_load()

# Set the paths for this notebook
set_paths(analysis_name="behavioral-shortened-removed", plots_name="human-llm-comparison-shortened-removed")
```

# Human-LLM accuracy analyses

## Data loading

Load data and set factors. Our columns are the following:
- **modality:** which modality was the stimulus presented as (video, audio, text, or model name) 
- **word_index:** the word index within the story
- **task:** which of the three stories was the data from (black, wheretheressmoke, howtodraw)

We create a contrast of modality such that we test the hypothesis that multimodal (video) > spoken (audio) > written (text).

```{r}

results_fn <- file.path(results_dir, "all-task_group-analyzed-behavior_window-size-00025_human-model-distributions-lemmatized.csv")

# Load data and convert columns to factors
df_results <- load_and_prepare_data(
  file_path=results_fn,
  factor_columns=c('modality', 'model_name', 'word_index', 'task', "Clause_Pos", "Clause_ID", "POS_spaCy")
)

# ADD A COLUMN THAT GROUPS HUMANS/MODELS
df_results <- df_results %>%
  mutate(model_type = case_when(
    model_name %in% c('roberta', 'electra', 'xlm-prophetnet') ~ "mlm",
    TRUE ~ "clm"  # Default case
  ))

# Set the levels for the modality column
df_results <- set_factor_levels(
  df=df_results,
  column_name="modality",
  levels=c("video", "audio", "text")
)

df_results <- create_column_contrast(
  df=df_results, 
  column_name="modality",
  n_poly=2
)

colnames(df_results)
```

## Analysis 1A - Human-LLM accuracy difference

```{r}

accuracy_diff_formula <- 'human_model_accuracy_diff ~  modality * model_entropy + (1|model_name) + (1|word_index:task)'

# Fit the linear mixed-effects model
model <- lmer(
  accuracy_diff_formula, 
  data=df_results, 
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Plot model results

```{r}

preds <- data.frame(get_model_data(model, type='pred', terms = c("model_entropy", "modality")))

colors <- c('#F068A1', '#82C564', '#F7CD84')

ggplot(preds, aes(x=x, y=predicted, color=group)) +
  geom_point(
    data = df_results,
    aes(x=model_entropy, y=human_model_accuracy_diff, color=modality),
    stroke=1,
    alpha=0.035) +
  geom_line(size=1) +
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill=group), alpha=0.3, color=NA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  theme(
    aspect.ratio = 16/9,
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  scale_color_manual(values=colors) +
  scale_fill_manual(values=colors) +
  labs(
    x = "Model Entropy",
    y = "Accuracy Difference",
    title = "Difference of human and model accuracy")

# ggsave(file.path(plots_dir, 'human-llm_accuracy-difference-entropy_new.pdf'), device = "pdf")

```

## Analysis 1B - Prosody interaction with accuracy difference

```{r}

accuracy_diff_formula <- 'human_model_accuracy_diff ~ modality * prominence_mean_words5 * model_entropy + (1|model_name) + (1|word_index:task)'

# Fit the linear mixed-effects model
model <- lmer(
  accuracy_diff_formula, 
  data=df_results, 
)

tab_model(model, show.se=TRUE, show.stat=TRUE, show.r2=TRUE, show.df=TRUE)

# Display results of fixed effects
display_model_summary(model)

# Conduct pairwise comparisons between modalities
pairwise_comparisons <- emmeans(model, pairwise ~ modality)
display_pairwise_comparisons(pairwise_comparisons)

```

### Plot the interaction with prosody

```{r}

preds <- data.frame(get_model_data(model, type='pred', terms = c("prominence_mean_words5", "modality")))
colors <- c('#F068A1', '#82C564', '#F7CD84')

ggplot(preds, aes(x=x, y=predicted, color=group)) + 
  geom_point(
    data = df_results, 
    aes(x=prominence_mean_words5, y=human_model_accuracy_diff, color=modality), 
    stroke=1,
    alpha=0.035) +
  geom_line(size=1) + 
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill=group), alpha=0.3, color=NA) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  theme(
    aspect.ratio = 5/4,
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  scale_color_manual(values=colors) +
  scale_fill_manual(values=colors) +
  labs(
    x = "Prosodic Prominence",
    y = "Accuracy Difference",
    title = "Prosody drives difference of human and LLM accuracy ") 

# ggsave(file.path(plots_dir, 'human-llm_accuracy-difference-prosody-interaction_new.pdf'), device = "pdf")
```
